Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-11 Response.tex

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
I have now added a \emph{Markov models of codon evolution} subsection in the introductory chapter, which discusses the most popular codon model implementations. The first paragraph of this new section is reproduced below (references are not reproduced correctly using the Latex `copy-over' tools we use in this response document):

\begin{quote}
\myeditsvonea
\end{quote}

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
%
%PL: TODO: rewrite the final paragraph to reflect the dated-tip likelihood model: Essential, we now consider heights (instead of branch lengths) and introduce a new parameter, the rate of evolution r, which allows rescaling genetic distance into time. The task is now to estimate the heights and r from the sequence data, keeping the the heights proportional to their sampling time, with the constant of proportionality being our estimable r.
%
I added a subsection which describes the methods used in BEAST to accommodate time-stamped sequences as a source for calibrating trees, when such information is available. The first paragraph of this new section is reproduced below:

@@

% 5
\item {
{\it
\textbf{
P20, 2nd display: an equality sign is missing on the 2nd line.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 6
\item {
{\it
\textbf{
I would find it useful to see a brief discussion on the plausibility of (1.22).
}% END: textbf
}% END: it
}% END: item
\\
% FB: Does he mean plausibility from a biological perspective or statistical? It's a necessary condition for the existence of stationary dist.
% Pl: I see it as questioning the plausibility in a biological sense. We now already argue that it is an assumption taken for computational reasons, but at the end of that paragraph we can explicitly state 'Realistic substitution processes are however unlikely to be time-reversible and procedures have also been developed to compute phylogenetic likelihoods under non-reversible models (cite Boussau and Gouy, 2006, see below)'. ...perhaps add in something more from that paper.... and end with 'We note that we use such non-reversible models in our discrete phytogeographic applications (cite Edwards et al., Current Biology, 2011).
% http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.324.620&rep=rep1&type=pdf
Please see my changes which explain how abandining the assumptions of stationarity and homogeneity of stochastic process governing the evolution entails each other and which start with:

\begin{quote}
\myeditsvsixone
\end{quote}

Relaxing both of these assumptions leads to processes which are also non-reversible, and addition to explaining the implications of the time-reversibility assumption from biological perspective, we also briefly introduce some of the recent attempts to relax this assumption, both in the Bayesian phylogeographic framework and more broadly in the context of phylogenetic applications:

### 2015-02-11 chapters/chapter01.tex
merge

Figure~\ref{fig:bayes1} shows the resulting posterior distribution along with the prior and the scaled likelihood.
The likelihood of the binomial model used in this example is simply:

@@

Phylogeography is a research direction aimed at connecting evolutionary processes, which happen over time, with the processes of spatial dispersal into a joint spatio-temporal dynamic that can be used to track pathogen spread.
%GB: I don't like the use of 'reciprocal' here ...
% FB: could also use complementary
Historically, phylogeographic inference in molecular epidemiology ignored the correlated nature of these two processes. 
The evolutionary relationships represented by the tree were inferred were first estimated from the sequence data, and then the ancestral locations generated by the process of the spatial diffusion were reconstructed from the spatial data, conditioning on the inferred tree topology.
\citet{Lemey2009} first considered the Bayesian reconstruction that allows for joint inference of both a time-measured evolutionary history and the parameters of the spatial and temporal process that shapes the spread and evolution of 
% e.g. 
rapidly evolving viruses.
This approach does not require fixing the tree topology, the branch lengths nor the Markov model parameters, and allows for integration over all possible values, requiring only for a specification of a prior distributions for the parameters that we aim to infer.

>>>>

\myedit{sveleven}{
Figure~\ref{fig:bayes1} shows the resulting posterior distribution along with the prior and the scaled likelihood.
The likelihood of the binomial model used in this example is simply:
}

@@

Phylogeography is a research direction aimed at connecting evolutionary processes, which happen over time, with the processes of spatial dispersal into a joint spatio-temporal dynamic that can be used to track pathogen spread.
Historically, phylogeographic inference in molecular epidemiology ignored the complementary character of these two processes. 
The evolutionary relationship represented by the tree was inferred omitting the geographical information, and then the ancestral locations generated by the process of the spatial diffusion were reconstructed, conditioning on the inferred tree topology.
\citet{Lemey2009} first considered the Bayesian reconstruction that allows for joint inference of both a time-measured evolutionary history and the parameters of the spatial and temporal process that shapes the spread and evolution of 
% e.g. 
rapidly evolving viruses.
This approach does not require fixing the tree topology, the branch lengths nor the Markov model parameters, and allows for integration over all possible values, requiring only for a specification of a prior distributions for the parameters that we aim to infer.
